---
title: "Google Play Store App Ratings, Installs and Pricing Analysis"
author:
  - name: "Hanwen Liu"
  - name: "Pratyush Kumar"
  - name: "Chenuli Muthukuda Arachchige"
  - name: "Shenyu He"
format:
  pdf:
    toc: true
    toc-title: "Table of Contents"
    include-in-header:
      text: |
        \usepackage{float}
bibliography: references.bib
nocite: '@*'
execute:
  echo: false
  warning: false
  message: false
---

## Executive Summary

This report investigates the key factors that influence mobile app success on the Google Play Store, focusing on ratings, install counts, pricing strategy, and app category. Using a dataset of over 10,000 apps combined with user review sentiment data, we find that free apps consistently attract higher install volumes, while paid apps tend to maintain marginally higher average ratings. Category emerges as a significant differentiator, with productivity and health apps outperforming entertainment categories in user satisfaction. Based on these findings, we recommend that developers prioritise free-with-monetisation models and invest in user experience within high-engagement categories.

## Introduction

The Google Play Store hosts over three million applications, making it one of the most competitive digital marketplaces in the world. Understanding what drives app success is critical for developers, investors, and product managers seeking to maximise reach and user satisfaction. Prior research suggests that app ratings serve as a primary signal of quality, directly influencing download decisions. However, the relationship between pricing, category, and performance remains incompletely understood. This study addresses the question: what factors influence mobile app success in terms of ratings and installs, and how do pricing and category affect performance? We analyse the Google Play Store dataset comprising 10,840 apps across 33 categories, supplemented by over 64,000 user reviews with sentiment scores. Our methodology combines descriptive statistics, visualisation, and comparative analysis across app types and categories. The findings are intended to provide actionable insights for app development strategy. This report is structured as follows: methodology, results, discussion, and conclusions.

## Methodology

The data for this study were sourced from the Google Play Store Apps dataset available on Kaggle, originally scraped using web scraping techniques. The dataset comprises googleplaystore.csv, containing 10,841 app records across 13 variables including app name, category, rating, reviews, installs, type (free or paid), and price.

Data cleaning was performed in R using the `tidyverse` package. Records with missing ratings were removed, and the `Installs` column was converted from character format (e.g., "1,000+") to numeric values by stripping commas and plus signs. The `Price` column was similarly cleaned by removing the dollar sign. Duplicate app entries were removed, retaining the most-reviewed version. After cleaning, 8,196 unique apps remained for analysis.

@tbl-summary presents a summary of the key numeric variables in the cleaned dataset, providing an overview of the central tendency and spread of ratings, reviews, and install counts.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics of key variables in the cleaned Google Play Store dataset"

library(tidyverse)
library(knitr)

apps <- read_csv("data/googleplaystore.csv", show_col_types = FALSE)

apps_clean <- apps |>
  filter(!is.na(Rating)) |>
  filter(Rating >= 1, Rating <= 5) |>
  mutate(
    Installs = as.numeric(gsub("[+,]", "", Installs)),
    Price = as.numeric(gsub("\\$", "", Price)),
    Reviews = as.numeric(Reviews)
  ) |>
  distinct(App, .keep_all = TRUE)

tibble(
  Variable = c("Rating", "Reviews", "Installs"),
  Mean = c(mean(apps_clean$Rating, na.rm = TRUE),
           mean(apps_clean$Reviews, na.rm = TRUE),
           mean(apps_clean$Installs, na.rm = TRUE)),
  Median = c(median(apps_clean$Rating, na.rm = TRUE),
             median(apps_clean$Reviews, na.rm = TRUE),
             median(apps_clean$Installs, na.rm = TRUE)),
  SD = c(sd(apps_clean$Rating, na.rm = TRUE),
         sd(apps_clean$Reviews, na.rm = TRUE),
         sd(apps_clean$Installs, na.rm = TRUE)),
  Min = c(min(apps_clean$Rating, na.rm = TRUE),
          min(apps_clean$Reviews, na.rm = TRUE),
          min(apps_clean$Installs, na.rm = TRUE)),
  Max = c(max(apps_clean$Rating, na.rm = TRUE),
          max(apps_clean$Reviews, na.rm = TRUE),
          max(apps_clean$Installs, na.rm = TRUE))
) |>
  mutate(across(where(is.numeric), ~ format(round(.x, 2), big.mark = ","))) |>
  kable()
```

To examine the relationship between pricing strategy and app performance, we compared the average ratings of free and paid apps across the ten most common categories. @fig-rating-by-type visualises this comparison using a grouped bar chart.

```{r}
#| label: fig-rating-by-type
#| fig-cap: "Average app rating by category and pricing type (free vs paid) for the top 10 categories"
#| fig-width: 7
#| fig-height: 3.5
#| fig-pos: "H"

top_categories <- apps_clean |>
  count(Category, sort = TRUE) |>
  slice_head(n = 10) |>
  pull(Category)

apps_clean |>
  filter(Category %in% top_categories, Type %in% c("Free", "Paid")) |>
  group_by(Category, Type) |>
  summarise(Avg_Rating = mean(Rating), .groups = "drop") |>
  ggplot(aes(x = reorder(Category, Avg_Rating), y = Avg_Rating, fill = Type)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip(ylim = c(3.5, 4.6)) +
  labs(
    x = "Category",
    y = "Average Rating",
    fill = "App Type"
  ) +
  theme_minimal(base_size = 11) +
  scale_fill_manual(values = c("Free" = "#4A90D9", "Paid" = "#D94A4A"))
```

Descriptive statistics and visualisations were used throughout the analysis. All computations were conducted in R (version 4.4), with `ggplot2` for plotting and `knitr` for table formatting, ensuring full reproducibility.



## Results

@fig-installs-by-type reveals a stark contrast in install volumes between free and paid apps. Free apps achieve a considerably higher median install count, with a wide interquartile range indicating that top-performing free apps attract massive user bases. Paid apps display a notably lower median with a more tightly compressed distribution, suggesting most paid apps accumulate relatively few installs. A low outlier further highlights that some paid apps struggle to even gain meaningful traction. These findings confirm that pricing strategy is a fundamental driver of app reach, with free apps commanding exponentially greater user acquisition.

```{r}
#| label: fig-installs-by-type
#| fig-cap: "Install volume distribution by app pricing type (free vs paid) on a log scale"
#| fig-width: 7
#| fig-height: 4
#| fig-pos: "H"

# Boxplot comparing install distributions for free vs paid apps

apps_clean |>
  filter(Type %in% c("Free", "Paid"), Installs > 0) |>
  ggplot(aes(x = Type, y = Installs, fill = Type)) +
  geom_boxplot(outlier.alpha = 0.2, width = 0.5) +
  scale_y_log10(labels = scales::comma) +
  labs(
    x = "App Type",
    y = "Number of Installs (log scale)"
  ) +
  theme_minimal(base_size = 11) +
  scale_fill_manual(values = c("Free" = "#4A90D9", "Paid" = "#D94A4A")) +
  theme(legend.position = "none")
```



```{r}
#| label: tbl-category-rating
#| tbl-cap: "Comparison of free and paid app counts and average ratings across the top 10 categories"


apps_clean |>
  filter(Category %in% top_categories, Type %in% c("Free", "Paid")) |>
  group_by(Category, Type) |>
  summarise(
    Count = n(),
    Avg_Rating = round(mean(Rating, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = Type,
    values_from = c(Count, Avg_Rating),
    names_glue = "{Type}_{.value}"
  ) |>
  rename(
    "Category" = Category,
    "Free Apps" = Free_Count,
    "Paid Apps" = Paid_Count,
    "Avg Rating (Free Apps)" = Free_Avg_Rating,
    "Avg Rating (Paid Apps)" = Paid_Avg_Rating
  ) |>
  arrange(desc(`Free Apps`)) |>
  kable()
```

@tbl-category-rating presents free and paid app counts alongside average ratings across the top 10 categories. Free apps dominate every category in volume, with Family and Game recording the highest counts. Personalization and Game emerge as the highest-rated categories overall, reflecting strong user engagement in entertainment-driven apps. Notably, paid apps consistently achieve marginally higher average ratings than free apps across most categories, with Personalization and Game showing the most pronounced differences. Finance, however, is a notable exception where free apps outperform. These findings confirm that free apps dominate in quantity, but paid apps deliver slightly higher user satisfaction across most categories.


## Discussion
The results show that pricing strategy is an important factor in app performance. Free apps receive much higher install volumes than paid apps, which means that users are more likely to download apps when there is no upfront cost. This makes free apps more effective for reaching a large audience.
However, paid apps often have slightly higher average ratings across most top categories. This may suggest that paid users are more engaged or that paid apps provide more specialised value. Category also matters, as Personalization and Game apps show strong average ratings, while Finance is an exception where free apps perform better than paid apps.

## Conclusion
In conclusion, the analysis shows that mobile app success depends on both pricing strategy and app category. Free apps are more successful in generating installs, while paid apps tend to achieve slightly higher average ratings in many categories. This means that install volume and user satisfaction represent different aspects of app performance.That means a free pricing model is more effective for maximising user reach, especially in competitive categories such as Family and Game. However, paid apps can still perform well when they provide strong value and maintain high user satisfaction. Therefore, software developers shouldn't evaluate app success using only downloads or ratings. A more complete assessment should consider installs, ratings, category, and pricing strategy together.

## Recommendation
Developers who aim to increase downloads and expand their market reach should consider using a free or freemium pricing model. Since free apps achieve substantially higher install volumes, this approach can help attract more users and build a larger audience. Developers can then use alternative monetisation strategies such as in-app purchases, subscriptions, or advertising.

For paid apps, developers should focus on delivering clear value and a high-quality user experience. Since paid apps generally have lower install volumes, strong ratings are important for building user trust and justifying the upfront cost. Paid app developers should clearly communicate the app’s benefits through descriptions, screenshots, reviews, and feature lists.

## References
