


## RESULTS 

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(knitr)

# Load the dataset

apps <- read_csv("data/googleplaystore.csv", show_col_types = FALSE)
```


```{r}
#| echo: false
#| message: false
#| warning: false

# Cleaning derived from from methodolgy section to make result section plot (delete after merging)

apps <- read_csv("data/googleplaystore.csv", show_col_types = FALSE)

apps_clean <- apps |>
  filter(!is.na(Rating)) |>
  filter(Rating >= 1, Rating <= 5) |>
  mutate(
    Installs = as.numeric(gsub("[+,]", "", Installs)),
    Price = as.numeric(gsub("\\$", "", Price)),
    Reviews = as.numeric(Reviews)
  ) |>
  distinct(App, .keep_all = TRUE)

```

```{r}
#| label: fig-installs-by-type
#| fig-cap: "Distribution of app installs by pricing type (free vs paid) on a log scale, highlighting the substantially higher install volumes of free apps"
#| fig-width: 7
#| fig-height: 4
#| warning: false
#| message: false

# Boxplot comparing install distributions for free vs paid apps

apps_clean |>
  filter(Type %in% c("Free", "Paid"), Installs > 0) |>
  ggplot(aes(x = Type, y = Installs, fill = Type)) +
  geom_boxplot(outlier.alpha = 0.2, width = 0.5) +
  scale_y_log10(labels = scales::comma) +
  labs(
    x = "App Type",
    y = "Number of Installs (log scale)"
  ) +
  theme_minimal(base_size = 11) +
  scale_fill_manual(values = c("Free" = "#4A90D9", "Paid" = "#D94A4A")) +
  theme(legend.position = "none")
```


@fig-installs-by-type reveals a stark contrast in install volumes between free and paid apps. Free apps achieve a considerably higher median install count, with a wide interquartile range indicating that top-performing free apps attract massive user bases. Paid apps, by contrast, display a notably lower median with a tightly compressed distribution, suggesting that the majority of paid apps accumulate relatively few installs. The presence of a low outlier further highlights that some paid apps struggle to gain any meaningful traction. This pattern directly addresses the research question: pricing strategy is a fundamental driver of app reach, with free apps commanding exponentially greater user acquisition than their paid counterparts.


