---
title: "Google Play Store App Ratings, Installs and Pricing Analysis"
subtitle: "ETC5513 Assignment 3 — Group 16"
author: "Hanwen Liu, Pratyush Kumar, Chenuli Muthukuda Arachchige, Shenyu He"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    incremental: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

## Problem Introduction

- The Google Play Store hosts millions of apps, making it one of the most competitive digital marketplaces in the world.

- For developers, understanding **what drives app success** is critical to maximising reach and user satisfaction.

- Two key success metrics: **user ratings** and **install counts**.

::: fragment
**Research question:** What factors influence mobile app success in terms of ratings and installs, and how do pricing and category affect performance?
:::

## Dataset Description

```{r}
library(tidyverse)
apps <- read_csv("data/googleplaystore.csv", show_col_types = FALSE)

apps_clean <- apps |>
  filter(!is.na(Rating), Rating >= 1, Rating <= 5) |>
  mutate(
    Installs = as.numeric(gsub("[+,]", "", Installs)),
    Price = as.numeric(gsub("\\$", "", Price))
  ) |>
  distinct(App, .keep_all = TRUE)
```

- **Source:** Google Play Store Apps dataset (Kaggle), collected via web scraping

- **`googleplaystore.csv`** — 10,841 raw records, reduced to **`r format(nrow(apps_clean), big.mark = ",")` unique apps** after cleaning, across **`r length(unique(apps_clean$Category))` categories**

  - Key variables: `Rating`, `Installs`, `Type` (Free/Paid), `Price`, `Category`

- **`googleplaystore_user_reviews.csv`** — over 64,000 user reviews with sentiment scores

::: fragment
Records with missing ratings and duplicate apps were removed to ensure data quality.
:::

## Methodology — Data Cleaning

Cleaning was performed in R using the `tidyverse` package:

- Removed records with **missing ratings**
- Converted `Installs` from text (e.g. "1,000+") to **numeric** values
- Cleaned `Price` (removed the `$` symbol)
- Removed **duplicate apps**, retaining the first occurrence

::: fragment
**Result: `r format(nrow(apps_clean), big.mark = ",")` unique apps** retained for analysis.
:::

## Methodology — Summary Statistics

```{r}

library(knitr)
library(kableExtra)
apps_stats <- apps_clean |>
  mutate(Reviews = as.numeric(Reviews))

tibble(
  Variable = c("Rating", "Reviews", "Installs"),
  Mean = c(mean(apps_stats$Rating, na.rm = TRUE),
           mean(apps_stats$Reviews, na.rm = TRUE),
           mean(apps_stats$Installs, na.rm = TRUE)),
  Median = c(median(apps_stats$Rating, na.rm = TRUE),
             median(apps_stats$Reviews, na.rm = TRUE),
             median(apps_stats$Installs, na.rm = TRUE)),
  SD = c(sd(apps_stats$Rating, na.rm = TRUE),
         sd(apps_stats$Reviews, na.rm = TRUE),
         sd(apps_stats$Installs, na.rm = TRUE))
) |>
  mutate(across(where(is.numeric), ~ format(round(.x, 2), big.mark = ","))) |>
  kable()
```

- Ratings are consistently high (mean **4.17**, low spread)
- Reviews & installs are **heavily skewed** — a few apps dominate the averages

## Methodology — Dataset Composition

```{r}
#| fig-width: 8
#| fig-height: 3.8

top_categories <- apps_clean |>
  count(Category, sort = TRUE) |>
  slice_head(n = 10) |>
  pull(Category)

apps_clean |>
  filter(Category %in% top_categories, Type %in% c("Free", "Paid")) |>
  count(Category, Type) |>
  ggplot(aes(x = reorder(Category, n), y = n, fill = Type)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  labs(x = "Category", y = "Number of Apps", fill = "App Type") +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("Free" = "#4A90D9", "Paid" = "#D94A4A"))
```

- Free apps **vastly outnumber** paid apps in every category
- Family, Game and Tools are the largest categories

## Free vs Paid: Install Distribution {.smaller}

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 5
#| fig-height: 4
#| out-width: "100%"

apps_clean |>
  filter(Type %in% c("Free", "Paid"), Installs > 0) |>
  ggplot(aes(x = Type, y = Installs, fill = Type)) +
  geom_boxplot(outlier.alpha = 0.2, width = 0.5) +
  scale_y_log10(labels = scales::comma) +
  labs(
    x = "App Type",
    y = "Number of Installs (log scale)"
  ) +
  theme_minimal(base_size = 11) +
  scale_fill_manual(values = c("Free" = "#4A90D9", "Paid" = "#D94A4A")) +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
::: incremental

- Free apps achieve substantially higher median install counts
- Paid apps cluster at lower installs with a compressed distribution
- Some paid apps struggle to gain meaningful traction
- Pricing strategy is a fundamental driver of app reach

:::
::::
::::::

## Results: Free vs Paid by Category {.smaller}

::: {style="font-size: 0.55em;"}
```{r}
apps_clean |>
  filter(Category %in% top_categories, Type %in% c("Free", "Paid")) |>
  group_by(Category, Type) |>
  summarise(
    Count = n(),
    Avg_Rating = round(mean(Rating, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = Type,
    values_from = c(Count, Avg_Rating),
    names_glue = "{Type}_{.value}"
  ) |>
  rename(
    "Category" = Category,
    "Free Apps" = Free_Count,
    "Paid Apps" = Paid_Count,
    "Avg Rating (Free Apps)" = Free_Avg_Rating,
    "Avg Rating (Paid Apps)" = Paid_Avg_Rating
  ) |>
  arrange(desc(`Free Apps`)) |>
  kable()
```
:::

::: incremental

- Family and Game dominate in app volume across all categories
- Personalization and Game emerge as the highest-rated categories
- Paid apps achieve marginally higher ratings than free apps overall
- Finance is a notable exception where free apps outperform paid
:::


## Discussion

::: incremental
- Pricing strategy shapes app performance, but its effect depends on the success metric.
- Free apps maximise reach — users download more readily with no upfront cost.
- Paid apps tend to earn slightly higher ratings, suggesting more intentional, value-focused users.
- Category matters too: Personalization and Game show strong engagement, while Finance bucks the trend with free apps outperforming paid.
:::

## Conclusion and Recommendation

::: incremental
- App success hinges on pricing, category, install volume, and ratings.
- Free apps drive more installs, especially in competitive categories like Family and Game.
- Paid apps can still thrive when they deliver clear value and a high-quality experience.
- For more downloads: go free or freemium, monetising via in-app purchases, subscriptions, or ads.
- Paid developers should build trust — communicate value through descriptions, screenshots, and reviews.
:::
